Token-level author diarization using clustering of stylistic contexts

### 2017-06-03 paper/paper.tex
no comment

% details for each component
\paragraph{Tokenization.} As a preprocessing step, we tokenize each document using \texttt{NLTK}\footnote{\url{http://www.nltk.org}} by \citet{bird-2009} to obtain a sequence of tokens $D$. We also perform part-of-speech tagging with the same tool, to speed up basic feature extraction which we describe below. We didn't use other preprocessing techniques such as lemmatization, stemming and stop word removal because they would take away a lot of stylometric data from text \cite{stamatatos-2009a}. The final output from the tokenization step was a finite sequence $\{(t_i, o_i, l(t_i), \mathit{POS}_i)\}_{i=1}^n$ where $o_i$ is the offset of token $t_i$, $l(t_i)$ its length in characters and $\mathit{POS}_i$ its POS tag.

\paragraph{Basic feature extraction.} 
We defined the stylistic context of a token $t_i$ as a set of tokens $\{t_k\}_{k=i-cs}^{i-1} \cup \{t_k\}_{k=i+1}^{i+cs}$ where $cs$ is a context size. From each context we extracted the most used stylometric features which we found in previous work. The features that we have considered in our models are:
\begin{itemize}
	\item \emph{Character tri-grams}. Frequencies of $n$-grams on character level have been very useful in quantifying the writing style \citep{stamatatos-2009b}. They are able to capture lexical and contextual information, use of punctuation and errors which can be an author's "fingerprint". This feature is also tolerant to noise. Based on work by \citet{stamatatos-2009a} and \citet{rahman-2015}, we choose $n=3$. Maximal dimension of this feature vector was set to $200$.
	\item \emph{Stop words}. According to \citet{stamatatos-2009a}, sometimes also called \emph{function words}, these are the most common used topic-independent words in text, such as articles, prepositions, pronouns and others. They are used unconsciously and found to be one of the most discriminative features in authorship attribution since they represent pure stylistic choices of authors' \citep{burrows-1987,argamon-2005}. We used frequencies of $156$ English stop words available in \texttt{NLTK}.
	\item \emph{Special characters}. We used counts of all character sequences which satisfied a regular expression defined as \verb/[^\\w]{1,4}/. Although character $n$-grams can catch the use of those character sequences, we wanted to have a distinct feature for that purpose since \citet{koppel-2009} mentioned that authors can have different punctuation habits.
	\item \emph{POS tag counts}. This is syntactic feature which \citet{koppel-2009} and \citet{stamatatos-2009a} also identified as a discriminative one in authorship analysis and it was used by \citet{kuznetsov-2016}. We used all $12$ tags from the universal tag set available in \texttt{NLTK}.
	\item \emph{Average token length}. Used by \citet{kuznetsov-2016}, \citet{sittar-2016},  \citet{brooke-2012} and \citet{stein-2011}. \citet{koppel-2009} characterized this feature as a complexity measure.
	\item \emph{Bag of Words}. Bag of words text representation more captures content, rather than style \citep{stamatatos-2009a}. We included this feature because it boosted performance of our initial testing, even without words been previously stemmed or lemmatized (stop words were excluded). Counts of a maximum of 100 unigrams were used.
	\item \emph{Type-token ratio}. We wanted to use that feature to measure the vocabulary richness in a token's context, but after we finished with performance evaluations we realized that there was a bug in implementation and its impact on final results was unknown to us. The feature should be calculated as the ratio of vocabulary size and total number of tokens of the text \citep{stamatatos-2009a}.
\end{itemize}

@@

\paragraph{Clustering.}
The final step in our approach is clustering of obtained feature vectors which represent stylistic contexts of tokens. The number of clusters is equal to the number of authors in tasks \emph{a} and \emph{b}, but in Task \emph{c} it is unknown. We tested various clustering algorithms available in \texttt{scikit-learn}\footnote{\url{http://scikit-learn.org}} Python toolkit by \citet{scikit-learn} in all tasks. In Model A for tasks  \emph{a} and \emph{b} we propose hierarchical agglomerative clustering, and for Task \emph{c} DBSCAN algorithm which estimates the number of clusters automatically. In Model B we use k-means for clustering in all tasks, and tried predicting the number of clusters based on the elbow rule. This is described in more detail in section \ref{sec:experimental-results}

@@

We performed initial testing of usefulness of all features except for the accidentally wrongly implemented type-token ratio which was added afterwards. For testing purposes we used the implementation of the k-means algorithm from \texttt{scikit-learn} with the default settings and all documents from the dataset for Task \emph{b}. As we were adding feature by feature, BCubed precision was becoming higher so we decided to use all features for Model A in all tasks. We don't know the exact impact of the wrongly implemented type-token ratio feature. Instead of the context's vocabulary size, we divided the size of document's vocabulary with the context size so this feature had the same value in all vectors.

@@

In every task we used the same tokenization procedure as described in section \ref{sec:the-proposed-approach} Model A covers all three tasks and has several hyperparameters for each of them: context size, clustering algorithm and clustering algorithm's hyperparameters. Model B uses a smaller set of features and trainable feature transformations with most of the hyperparameters equal among the tasks.

\paragraph{Model A.} 
Sliding widow context size was determined manually for tasks \emph{a} and \emph{b}, in a similar way as feature selection. We used a whole dataset and default implementation of k-means from \texttt{scikit-learn} and tested different values. The best result for Task \emph{a} was $cs_a=16$ and for Task \emph{b} $cs_b=140$. For Task \emph{c}, we left $cs_c=140$ as for Task \emph{b}. 
%This hyperparameter probably depends on document's length to a certain extent. If we compare that sizes with the average document lengths in tokens across datasets, we can see that there is a correspondence ($1679, 3767, 3298$ vs. $16, 140, 140$). However, they are probably biased to a certain extent.

@@

Since the model with the best achieved Bcubed $F_1$-score ($0.61$) is not of practical use in real life applications because the number of clusters is always wrongly estimated, we decided to repeat the grid search in Task \emph{c} for DBSCAN, but this time we used the Calinski-Harabaz index defined by \citet{calinski-1974} as evaluation measure since it does not require the real number of clusters for clustering evaluation. Finally, from all results on training set we chose (\texttt{eps=0.23, min\_samples=64, metric="cosine"}) for the final hyperparameters. This wasn't the best combination, but it produced relatively acceptable amount of noisy elements per document, and performed quite fast but again din't estimate the correct number of clusters in most cases. If some samples weren't clustered, we assigned them the majority label from neighborhood (context) of size $10$ ($5$ on each side), or the majority label from all samples if samples in the context weren't clustered.

\paragraph{Model B.} 
For Model B, we have decided to use a smaller set of basic features: character tri-gram counts (for $120$ tri-grams), stop word counts, average token length and bag-of-words (for $120$ words). The length of the sliding window is set to $120$ tokens. Model B uses k-means clustering. We have chosen two trainable feature transformations: linear and elementwise linear (weighting). For both of the transformations $5\times 10^{-4}$ was chosen for the learning rate. All the hyperparameters are selected manually by experimenting on the development dataset, which is approximately $70\%$ of the available dataset. $70\%$ of the development set was used for training basic feature extractors and the transformation. Considering the results on the validation part of the development set, not very large numbers of iterations seem to be more beneficial. A number of $40$ iterations was chosen for the linear transformation, and $20$ for the elementwise linear transformation. For the output dimension of the linear transformation we have chosen $n_\mathrm{t}=40$, though not much different results seem to be achievable using smaller numbers like $20$, making training and clustering faster.

@@

The results of evaluation for the intrinsic plagiarism problem are shown in Table \ref{tab:results-task-a}. For author diarization with known and unknown numbers of authors, the results are shown in Table \ref{tab:results-task-b-c}. It should be noted that the evaluated Model B uses a fixed number of clusters $k=2$ for task \emph{c} and that the results are actually not very good. By trying a more honest approach with predicting the number of clusters the BCubed $F_1$-score drops below $0.5$. The results of Stohastic dummy baseline for all tasks are obtained over the 50 runs, as well as for Simple baseline in task \emph{c}.

>>>>

% details for each component
\paragraph{Tokenization.} As a preprocessing step, we tokenize each document using NLTK\footnote{\url{http://www.nltk.org}} by \citet{bird-2009} to obtain a sequence of tokens $D$. We also perform part-of-speech tagging with the same tool, to speed up basic feature extraction which we describe below. We didn't use other preprocessing techniques such as lemmatization, stemming and stop word removal because they would take away a lot of stylometric data from text \cite{stamatatos-2009a}. The final output from the tokenization step was a finite sequence $\{(t_i, o_i, l(t_i), \mathit{POS}_i)\}_{i=1}^n$ where $o_i$ is the offset of token $t_i$, $l(t_i)$ its length in characters and $\mathit{POS}_i$ its POS tag.

\paragraph{Basic feature extraction.} 
We defined the stylistic context of a token $t_i$ as a set of tokens $\{t_k\}_{k=i-cs}^{i-1} \cup \{t_k\}_{k=i+1}^{i+cs}$ where $cs$ is a context size. From each context we extracted the most used stylometric features which we found in previous work. The features that we have considered in our models are:
\begin{itemize}
	\item \emph{Character tri-grams}. Frequencies of $n$-grams on character level have been very useful in quantifying the writing style \citep{stamatatos-2009b}. They are able to capture lexical and contextual information, use of punctuation and errors which can be an author's "fingerprint". This feature is also tolerant to noise. Based on work by \citet{stamatatos-2009a} and \citet{rahman-2015}, we choose $n=3$. Maximal dimension of this feature vector was set to $200$.
	\item \emph{Stop words}. According to \citet{stamatatos-2009a}, sometimes also called \emph{function words}, these are the most common used topic-independent words in text, such as articles, prepositions, pronouns and others. They are used unconsciously and found to be one of the most discriminative features in authorship attribution since they represent pure stylistic choices of authors' \citep{burrows-1987,argamon-2005}. We used frequencies of $156$ English stop words available in NLTK.
	\item \emph{Special characters}. We used counts of all character sequences which satisfied a regular expression defined as \verb/[^\\w]{1,4}/. Although character $n$-grams can catch the use of those character sequences, we wanted to have a distinct feature for that purpose since \citet{koppel-2009} mentioned that authors can have different punctuation habits.
	\item \emph{POS tag counts}. This is syntactic feature which \citet{koppel-2009} and \citet{stamatatos-2009a} also identified as a discriminative one in authorship analysis and it was used by \citet{kuznetsov-2016}. We used all $12$ tags from the universal tag set available in NLTK.
	\item \emph{Average token length}. Used by \citet{kuznetsov-2016}, \citet{sittar-2016},  \citet{brooke-2012} and \citet{stein-2011}. \citet{koppel-2009} characterized this feature as a complexity measure.
	\item \emph{Bag of Words}. Bag of words text representation more captures content, rather than style \citep{stamatatos-2009a}. We included this feature because it boosted performance of our initial testing, even without words been previously stemmed or lemmatized (stop words were excluded). Counts of a maximum of 100 unigrams were used.
	\item \emph{Type-token ratio}. We wanted to use that feature to measure the vocabulary richness in a token's context, but after we finished with performance evaluations we realized that there was a bug in implementation and its impact on final results was unknown to us. The feature should be calculated as the ratio of vocabulary size and total number of tokens of the text \citep{stamatatos-2009a}.
\end{itemize}

@@

\paragraph{Clustering.}
The final step in our approach is clustering of obtained feature vectors which represent stylistic contexts of tokens. The number of clusters is equal to the number of authors in tasks \emph{a} and \emph{b}, but in Task \emph{c} it is unknown. We tested various clustering algorithms available in scikit-learn\footnote{\url{http://scikit-learn.org}} Python toolkit by \citet{scikit-learn} in all tasks. In Model A for tasks  \emph{a} and \emph{b} we propose hierarchical agglomerative clustering, and for Task \emph{c} DBSCAN algorithm which estimates the number of clusters automatically. In Model B we use k-means for clustering in all tasks, and tried predicting the number of clusters based on the elbow rule. This is described in more detail in section \ref{sec:experimental-results}

@@

We performed initial testing of usefulness of all features except for the accidentally wrongly implemented type-token ratio which was added afterwards. For testing purposes we used the implementation of the k-means algorithm from scikit-learn with the default settings and all documents from the dataset for Task \emph{b}. As we were adding feature by feature, BCubed precision was becoming higher so we decided to use all features for Model A in all tasks. We don't know the exact impact of the wrongly implemented type-token ratio feature. Instead of the context's vocabulary size, we divided the size of document's vocabulary with the context size so this feature had the same value in all vectors.

@@

In every task we used the same tokenization procedure as described in section \ref{sec:the-proposed-approach} Model A covers all three tasks and has several hyperparameters for each of them: context size, clustering algorithm and clustering algorithm's hyperparameters. Model B uses a smaller set of features and trainable feature transformations with most of the hyperparameters equal among the tasks. All the source code for our experiments is available in our repository\footnote{\url{https://github.com/Coolcumber/inpladesys/}}.

\paragraph{Model A.} 
Sliding widow context size was determined manually for tasks \emph{a} and \emph{b}, in a similar way as feature selection. We used a whole dataset and default implementation of k-means from scikit-learn and tested different values. The best result for Task \emph{a} was $\mathit{cs}_\mathrm{a}=16$ and for Task \emph{b} $\mathit{cs}_\mathrm{b}=140$. For Task \emph{c}, we left $\mathit{cs}_\mathrm{c}=140$ as for Task \emph{b}. 
%This hyperparameter probably depends on document's length to a certain extent. If we compare that sizes with the average document lengths in tokens across datasets, we can see that there is a correspondence ($1679, 3767, 3298$ vs. $16, 140, 140$). However, they are probably biased to a certain extent.

@@

Since the model with the best achieved Bcubed $F_1$-score ($0.61$) is not of practical use in real life applications because the number of clusters is always wrongly estimated, we decided to repeat the grid search in Task \emph{c} for DBSCAN, but this time we used the Calinski-Harabaz index defined by \citet{calinski-1974} as evaluation measure since it does not require the real number of clusters for clustering evaluation. Finally, from all results on training set we chose (\texttt{eps=0.23, min\_samples=64, metric="cosine"}) for the final hyperparameters. This wasn't the best combination, but it produced relatively acceptable amount of noisy elements per document, and performed quite fast but again didn't estimate the correct number of clusters in most cases. If some samples weren't clustered, we assigned them the majority label from neighborhood (context) of size $10$ ($5$ on each side), or the majority label from all samples if samples in the context weren't clustered.

\paragraph{Model B.} 
For Model B, we have decided to use a smaller set of basic features: character tri-gram counts (for $120$ tri-grams), stop word counts, average token length and bag-of-words (for $120$ words). The length of the sliding window is set to $120$ tokens. Model B uses k-means clustering. We have chosen two trainable feature transformations: linear and elementwise linear (weighting). For both of the transformations $5\times 10^{-4}$ was chosen for the learning rate. All the hyperparameters are selected manually by experimenting on the development dataset, which is approximately $70\%$ of the available dataset. $70\%$ of the development set was used for fitting basic feature extractors and learning the transformation. Considering the results on the validation part of the development set, not very large numbers of iterations seem to be more beneficial. A number of $40$ iterations was chosen for the linear transformation, and $20$ for the elementwise linear transformation. For the output dimension of the linear transformation we have chosen $n_\mathrm{t}=40$, though not much different results seem to be achievable using smaller numbers like $20$, making training and clustering faster.

@@

The results of evaluation for the intrinsic plagiarism problem are shown in Table \ref{tab:results-task-a}. For author diarization with known and unknown numbers of authors, the results are shown in Table \ref{tab:results-task-b-c}. It should be noted that the evaluated Model B uses a fixed number of clusters $k=2$ for task \emph{c} and that the results are actually not very good. By trying a more honest approach with predicting the number of clusters the BCubed $F_1$-score drops below $0.5$. The results of Stohastic dummy baseline for all tasks are obtained over the 50 runs, as well as for Simple baseline in Task \emph{c}.
